---
title: 'Estudo de Caso 01: Comparação do IMC médio de alunos do PPGEE-UFMG ao longo de dois semestres'
author: "Beatriz S. Campagnaro, Izabela C. S. Barbosa, Filipe C. Andrade e Stephanie F. Lemos"
date: "28 de Agosto de 2026"
output:
  pdf_document:
    fig_caption: yes
  html_document:
    df_print: paged
---

```{r setup, results='hide', warning=FALSE, echo=FALSE, message=FALSE}

# Definindo o mirror do CRAN
options(repos = c(CRAN = "[https://cloud.r-project.org](https://cloud.r-project.org)"))

# Instalando/carregando ggplot2
if (!require(ggplot2, quietly = TRUE)){
      install.packages("ggplot2")
      library(ggplot2)
}

# Instalando/carregando broom DIRETO DO CRAN
if (!require(broom, quietly = TRUE)){
      install.packages("broom")
      library(broom)
}

# Instalando/carregando GGally
if (!require(GGally, quietly = TRUE)){
      install.packages("GGally")
      library(GGally)
}

# Instalando/carregando dplyr
if (!require(dplyr, quietly = TRUE)){
      install.packages("dplyr")
      library(dplyr)
}
```

## Introdução TESTE TESTE

Nesse estudo será realizada a comparação do IMC (Índice de Massa Corporal) médio de duas populações de estudantes, sendo estas compostas por alunos de pós graduação em Engenharia Elétrica na UFMG (Universidade Federal de Minas Gerais) nos semestres 2016-2 e 2017-2.

### 1. Coleta e Tratamento de Dados

O prmeiro passo empregado foi a coletada de dados, na qual são coletados os dados disponibilizados por meio dos aquivos .CSV, em seguida esses dados foram tratados para ficarem no mesmo formato, deixando suas colunas da seguinte forma

ID, Peso, Altura, Genero, Curso, Peso e Idade

```{r configuracao_fontes}

# 1. Configuração dos arquivos e das colunas
# Cada semestre é definido por: caminho do arquivo, separador,
# mapeamento das colunas, filtro de curso e identificação do semestre.

config_2016 <- list(
  caminho      = "Dados/imc_20162.csv",
  separador    = ",",
  mapa_colunas = c(ID = "ID", Curso = "Course", Genero = "Gender",
                    Altura = "Height.m", Peso = "Weight.kg"),
  filtro_curso = "PPGEE",   # 2016 vem misturado com a graduação
  semestre     = "2016-2"
)

config_2017 <- list(
  caminho      = "Dados/imc_20172.csv",
  separador    = ";",
  mapa_colunas = c(Peso = "Weight.kg", Altura = "height.m",
                    Genero = "Sex", Idade = "Age.years"),
  filtro_curso = NULL,      # 2017 já vem só com PPGEE
  semestre     = "2017-2"
)
```

```{r carregar_dados}
# 2. Carregamento e padronização dos dados

carregar_dados <- function(config) {
  bruto <- read.csv(config$caminho,
    sep = config$separador,
    stringsAsFactors = FALSE)

  bruto <- rename(bruto, !!!config$mapa_colunas)

  # Criação das colunas ausentes no arquivo original
  if (!"Idade" %in% names(bruto)) bruto$Idade <- NA_real_
  if (!"Curso" %in% names(bruto)) bruto$Curso <- "PPGEE"
  if (!"ID" %in% names(bruto))    bruto$ID    <- seq_len(nrow(bruto))

  # Aplicação do filtro de curso quando necessário
  if (!is.null(config$filtro_curso)) {
    bruto <- filter(bruto, Curso == config$filtro_curso)
  }

  bruto$Semestre <- config$semestre
  select(bruto, Semestre, ID, Curso, Genero, Idade, Altura, Peso)
}

dados_2016 <- carregar_dados(config_2016)
dados_2017 <- carregar_dados(config_2017)
```

O arquivo de 2016 contém também alunos de graduação (curso ENGSIS) e por isso exige filtro adicional pelo curso PPGEE; o de 2017 já contém exclusivamente alunos de pós-graduação. 

Em seguida foi calculado o Índice de Massa Corporal (IMC) de cada participante por meio da equação:

$$IMC = \frac{Peso}{Altura^2}$$

sendo o peso dado em quilogramas (kg) e a altura em metros (m).

```{r calcular_imc}
# 3. Cálculo do Índice de Massa Corporal (IMC)
calcular_imc <- function(dados) {
  mutate(dados, IMC = Peso / (Altura^2))
}

dados_2016 <- calcular_imc(dados_2016)
dados_2017 <- calcular_imc(dados_2017)
```

Antes dos testes de hipótese, os dados foram verificados quanto à presença de valores ausentes e identificadores duplicados. Não foram aplicados filtros por faixa de altura ou peso, a fim de evitar a exclusão de observações potencialmente legítimas. A base de 2016-2 continha 28 observações (21 homens e 7 mulheres) e a de 2017-2, 25 (21 homens e 4 mulheres). Não foram identificados valores ausentes ou duplicidades, portanto, nenhuma observação foi excluída.

```{r checar_qualidade}
checar_qualidade <- function(dados) {
  # 4. Verificação da qualidade dos dados
  dados %>%
    group_by(Semestre) %>%
    summarise(
      n_total           = n(),
      n_ausentes_peso   = sum(is.na(Peso)),
      n_ausentes_altura = sum(is.na(Altura)),
      n_id_duplicado    = sum(duplicated(ID)),
      .groups = "drop"
    )
}
# 5. Consolidação das bases
dados <- bind_rows(dados_2016, dados_2017)
checar_qualidade(dados)
```
## Experimental design

A section detailing the experimental setup. This is the place where you will define your test hypotheses, e.g.: $$\begin{cases} H_0: \mu = 10&\\H_1: \mu<10\end{cases}$$

including the reasons behind your choices of the value for $H_0$ and the directionality (or not) of $H_1$.

This is also the place where you should discuss (whenever necessary) your definitions of minimally relevant effects ($\delta^*$), sample size calculations, choice of power and significance levels, and any other relevant information about specificities in your data collection procedures.

### Description of the data collection

Whenever needed, you can also include an (optional) subsection describing the actual data collection, how it was performed, any adaptations or unexpected events that may have occurred, etc. Subsections like this can also be used for the sample size calculations or any other aspect that requires a longer discussion.

## Exploratory Data Analysis

The first step is to load and preprocess the data. For instance,

```{r loaddata}
data(mtcars)
fc<-c(2,8:11)
for (i in 1:length(fc)){mtcars[,fc[i]]<-as.factor(mtcars[,fc[i]])}
levels(mtcars$am) <- c("Automatic","Manual")
```

To get an initial feel for the relationships between the relevant variables of your experiment it is frequently interesting to perform some preliminary (exploratory) analysis. This is frequently referred to as *getting a feel* of your data, and can suggest procedures (such as outlier investigation or data transformations) to experienced experimenters.

```{r pairs,fig.width=8,echo=TRUE,message=FALSE,fig.cap="Exploring the effect of car transmission on mpg values"}
library(GGally,quietly = T, warn.conflicts = F) # This is just me getting fancy.
                                                # There are much simpler ways ;-)
ggpairs(data=mtcars,columns=c(1,9),title="MPG by transmission type",
        upper=list(combo="box"),lower=list(combo="facethist"),
        diag=list(continuous="densityDiag",discrete="barDiag"))
```

Your preliminary analysis should be described together with the plots. In this example, two facts are immediately clear from the plots: first, **mpg** tends to correlate well with many of the other variables, most intensely with **drat** (positively) and **wt** (negatively). It is also clear that many of the variables are highly correlated (e.g., **wt** and **disp**). Second, it seems like manual transmission models present larger values of **mpg** than the automatic ones. In the next section a linear model will be fit to the data in order to investigate the significance and magnitude of this possible effect.

## Statistical Analysis

Your statistical analysis should come here. This is the place where you should fit your statistical model, get the results of your significance test, your effect size estimates and confidence intervals.

```{r fitmodel,results='hold'}
model<-aov(mpg~am*disp,data=mtcars)
summary(model)
```

### Checking Model Assumptions

The assumptions of your test should also be validated, and possible effects of violations should also be explored.

```{r resplots,fig.width=8,echo=TRUE,message=FALSE, fig.cap = "Residual plots for the anova model"}
par(mfrow=c(2,2), mai=.3*c(1,1,1,1))
plot(model,pch=16,lty=1,lwd=2)
```

### Conclusions and Recommendations

The discussion of your results, and the scientific/technical meaning of the effects detected, should be placed here. Always be sure to tie your results back to the original question of interest!

### Declaration of generative AI and AI-assisted technologies in the report preparation process.

During the preparation of this work the author(s) used [NAME OF TOOL / SERVICE] in order to [REASON]. After using this tool/service, the author(s) reviewed and edited the content as needed and take(s) full responsibility for the content of the published article.
